Generalized universal instability: Transient linear amplification and subcritical turbulence

### 2015-03-20 universalModeTurbulencePaper.tex
Updates for referee response

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\omegase}{\omega_{*}}
\newcommand{\nuee}{\nu_{ee}}
\newcommand{\energy}{\mathcal{E}}
\newcommand{\sgn}{\mathrm{sgn}}
\newcommand{\changed}[1]{\textcolor{red}{#1}}
%\newcommand{\changed}[1]{#1}
\newcommand{\transpose}{\mathrm{T}}

@@

We consider slab geometry with magnetic shear, so the magnetic field in Cartesian
coordinates $(x',y',z')$ is $\vect{B} = [\vect{e}_{z'} + (x' / L_s) \vect{e}_{y'} ]B$
for some constant $B$ and shear length $L_s$.  A field-aligned coordinate system
is introduced: $x = x', \;y = y' - x' z' / L_s, \; z = z'$,
so $\vect{B}\cdot\nabla x = \vect{B}\cdot\nabla y = 0$ and $\vect{B}\cdot\nabla z = B$.
In this geometry, we consider the nonlinear electrostatic
gyrokinetic equation \cite{FriemanChen},
dropping toroidal effects:
\begin{eqnarray}
\label{eq:gke}
\frac{\partial h_s}{\partial t}
+ v_{||} \frac{\partial h_s}{\partial z}
+ \frac{1}{B} \left[
\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial X_s} \frac{\partial h_s}{\partial Y_s}
-\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial Y_s} \frac{\partial h_s}{\partial X_s}
\right]
-\left< C_s\left\{h_s\right\} \right>_{\vect{R}_s} \\
=
\frac{q_s}{T_s} f_{Ms} \frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial t}
-\frac{f_{Ms}}{B L_n} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right],
\nonumber
\end{eqnarray}
together with the quasineutrality condition
\begin{equation}
-\frac{n e \Phi}{T_i} + \int d^3v \left<h_i\right>_{\vect{r}}
=
\frac{n e \Phi}{T_e} + \int d^3v \left<h_e\right>_{\vect{r}}.
\label{eq:qn}
\end{equation}
Here $s \in \left\{i,e\right\}$ subscripts denote species,
$\Phi(t,\vect{r})$ is the electrostatic potential, $h_s(t,\vect{R}_s,\energy_s,\mu)$ is the nonadiabatic distribution function,
$\energy_s = m_s v^2/(2 T_s)$, $\mu = v_{\bot}^2/(2B)$,
and $f_{Ms}$ is the leading-order Maxwellian.
The scale length of the equilibrium density $n$ is $L_n = -n/(dn/dx)$,
$q_i = e = -q_e$ is the proton charge,
$T_s$ is the species temperature,
\changed{$\eta_s=(d \ln T_s/dx)/(d \ln n/dx)$,}
$C_s$ is the collision operator,
and
$\left<\right>_{\vect{r}}$
and
$\left<\right>_{\vect{R}_s}$
denote gyroaverages at fixed particle position $\vect{r}=(x,y,z)$ and guiding-center position $\vect{R}_s=(X_s,Y_s,Z_s)$.
The distinction between $z$ and $Z_s$ may be neglected.
Expanding the fluctuating quantities in spatial Fourier modes as $\Phi = \sum_{k_x,k_y} \Phi_{k_x,k_y}(t,z) \exp(i k_x x + i k_y y)$
and $h_s = \sum_{k_x,k_y} h_{s,k_x,k_y}(t,z,\energy_s,\mu) \exp(i k_x X_s + i k_y Y_s)$,
we may replace the gyro-averages with Bessel functions:
$\left<\Phi_{k_x,k_y} \exp(i k_x x + i k_y y)\right>_{\vect{R}_s}
=J_{0s} \Phi_{k_x,k_y} \exp(i k_x X_s + i k_y Y_s)$
and
$\left<h_{s,k_x,k_y}\exp(i k_x X_s + i k_y Y_s)\right>_{\vect{r}} = J_{0s} h_{s,k_x,k_y}\exp(i k_x x + i k_y y)$
with $J_{0s} = J_0(k_{\perp} v_{\perp} / \Omega_s)$
and $\Omega_s = e B/m_s$.
The perpendicular wavenumber is given by
\begin{equation}
k_{\perp} = \sqrt{k_y^2 + (k_x - k_y z/L_s)^2}.
\label{eq:kperp}
\end{equation}
The system is uniquely specified by the following dimensionless parameters:
$L_s / L_n$, $\eta_i$, $\eta_e$, $T_e / T_i$, and collisionality $\nuee L_n / v_i$,
where $\nuee = \sqrt{2}\pi n e^4 \ln\Lambda/[(4\pi\epsilon_0)^2 m_e^{1/2} T_e^{3/2}]$ is the electron-electron collision frequency
and $v_i = \sqrt{2 T_i / m_i}$.

@@

One important property of the system (\ref{eq:gke})-(\ref{eq:qn})
is the free energy conservation equation \cite{Howes, AbelReview}
obtained from the $\sum_s \int d^3R_s \int d^3v (T_s h_s/f_{Ms}) (\ldots)$ moment of (\ref{eq:gke}):
\begin{equation}
\frac{d W}{d t} = S(h_i,h_e),
\label{eq:dWdt}
\end{equation}
where
\begin{equation}
\label{eq:W}
W = \sum_s \int d^3r \int d^3v \frac{T_s \, \delta\! f_s^2}{2 f_{Ms}}
\end{equation}
is the free energy, $\delta\! f_s = h_s - q_s \Phi f_{Ms}/T_s$ is the total departure of the distribution
function from a Maxwellian,
and
\begin{eqnarray}
\label{eq:freeEnergySources}
S(h_i,h_e)
&=&
\sum_s \int d^3r \int d^3v \frac{T_s h_s}{f_{Ms}}  C_s\{ h_s \}  \\
&&+\sum_s
\int d^3R_s \int d^3v \frac{h_s}{B} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
 \frac{T_s}{n_s}\frac{dn_s}{dx}
\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right]
\nonumber
\end{eqnarray}
is the net energy source.
(To derive (\ref{eq:dWdt})-(\ref{eq:freeEnergySources}), the equalities
$\int d^3 R_s \int d^3v \left< \ldots\right>_{\vect{R}_s}
= \int d^3 R_s \int d^3v (\ldots)
\approx \int d^3 r \int d^3v (\ldots)
= \int d^3 r \int d^3v \left< \ldots\right>_{\vect{r}}
$ are used.)
The first term on the right-hand side of (\ref{eq:freeEnergySources})
is negative-definite (by the $H$-theorem).
The second term has the form of the particle and heat fluxes multiplied by
the density and temperature gradients, and this term increases $W$ if the fluxes have the appropriate sign
to relax the gradients.
Thus, in a turbulent steady state where $W$ is statistically constant,
equations (\ref{eq:dWdt})-(\ref{eq:freeEnergySources}) indicate
that energy is injected by the equilibrium gradient(s)
and dissipated by collisions.
Note that the operation $\int d^3R_s$ used in the derivation of
(\ref{eq:dWdt}) annihilates the nonlinear term
in (\ref{eq:gke}), and so (\ref{eq:dWdt})-(\ref{eq:freeEnergySources})
apply to both the linear and nonlinear dynamics.

@@

Results for the three norms shown are quite similar,
though the curve for the $W$ norm appears slightly higher than the other curves
due to less decay in the initial period $t v_i / L_n <20$.
Indeed, we find
all results described or plotted for the remainder of this paper are quite similar
for all of these norms.
Henceforth we restrict our attention to the $W$ norm since it is well motivated
by the conservation equation (\ref{eq:dWdt}).

For the calculation in figure \ref{fig:transientAmplification}, the initial condition used
is the standard ``noise'' initial condition of gs2:
$g_s = h_s - \left< \Phi\right>_{\vect{R}_s} q_s f_{Ms} / T_s$ is set to a Maxwellian velocity distribution
times a random complex number at each grid point in $z$.
(The real and imaginary parts of these complex numbers are each taken from a uniform distribution on $[-1,1]$.)
While techniques exist to calculate the optimal perturbation
that gives maximum amplification \cite{FarrellIoannou, SquirePRL, SquireApJ}, such techniques
are numerically challenging for this kinetic model due to the high dimensionality
of the problem.
The noise initial condition is a convenient shortcut, ensuring energy is likely
to be given to a growing perturbation if one exists.
It has been shown in \cite{Camargo, SquireApJ} that for studies of transient linear
amplification, random initial conditions are a reasonable proxy
for a directly calculated optimal initial condition.

\begin{figure}[h!]
% Figure generated with m20141119_01_plotGS2TransientAmplificationForPaper_figureA.m
\includegraphics[width=3in]{m20141119_01_plotGS2TransientAmplificationForPaper_figureA_a.pdf}
% Figure generated with m20141121_01_plotGS2TransientAmplificationForPaper.m
\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_b.pdf}
\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_c.pdf}
\caption{(Color online) (a) Linear evolution for $L_s/L_n=300$, $\eta_i = \eta_e = 0$,
$\nuee = 0.05 v_i/L_n$, and $k_y \rho_i = 1$, illustrating transient amplification in three norms.
The initial conditions are detailed in the main text.
(b) The amount of transient amplification may be defined relative to $t=0$ or relative
to the minimum that occurs shortly thereafter.
(c) The potential structure along a field line for this example.
\label{fig:transientAmplification}}
\end{figure}

Figure \ref{fig:transientAmplification}.b shows the same computation, zooming in to the initial transient period.
Before the amplification begins around $t v_i/L_n=20$, there is an initial decay period due to strongly
damped modes. An amplification factor may be defined relative to either to this minimum amplitude,
or relative to the amplitude at $t=0$.
In both cases, we define the amplification factor to be 1
if there is no such transient amplification.

@@

\changed{
To demonstrate this principle across a range of parameters,
figure  \ref{fig:comparingTransientTo0Shear} demonstrates that the
correspondence between the finite-shear transients and zero-shear eigenmodes
is maintained throughout the evolution of the transient, for a range of
values of magetic shear.}
%This comparison is shown in figure \ref{fig:comparingTransientTo0Shear}.
In this figure, we display 25 independent simulations of linear transients,
varying $L_s/L_n$ between 20 and 300, and plotting the evolution of $\gamma = (2W(t))^{-1} dW/dt$ from the time
of peak growth rate to the time at which the energy begins to decay.
For all simulations shown, $k_y \rho_i = 2$ and $\nuee = 0.05 v_i/L_n$.
In each case, there is very close agreement between the instantaneous growth rate of the transient
and the `predicted' growth rate from the zero-shear dispersion relation.
Thus, we can conclude that the transiently growing structures are closely related to the zero-shear
eigenmodes.  
%Magnetic shear causes $k_{\perp}$ and $k_{||}$ to vary with time as the structure
%evolves (unlike the zero-shear case in which $k_{\perp}$ and $k_{||}$ are constants.)

\begin{figure}[h!]
% Data generated on Edison in /global/scratch2/sd/landrema/20150109-01-coderunner-edison-0shearUniversalInstability/vnewk_e_0p05
% Figure generated with m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.m
\includegraphics[width=4in]{m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.pdf}
\caption{(Color online)
Although a change from zero to small finite magnetic shear causes a fundamental change in the eigenvalues,
(which change from unstable to decaying),
the \emph{transient} finite-shear structures retain a similarity to the zero-shear modes.  For as shown here,
the instantaneous growth rate of the transient finite-shear structure is nearly identical
to the growth rate of a zero-shear eigenmode at the dominant $k_{\perp}$ and $k_{||}$ values
(\ref{eq:expectedKperp})-(\ref{eq:expectedKpar}).  Each trajectory shown is an independent
linear initial-value calculation, with $L_s/L_n$ varying from 20-300 between simulations.  The growth rate is plotted from
the time of maximum growth rate (at the $+$ symbols) until decay begins.
\label{fig:comparingTransientTo0Shear}}
\end{figure}

@@

@@

Figure \ref{fig:transientAmplificationVsKy} shows the dependence
of the amplification factor on $k_y$ at fixed $L_s/L_n = 200$.
Significant amplification can be observed for a range of $k_y$ surrounding
$1/\rho_i$.
Results are shown for two values of collisionality,
$\nuee L_n / v_i = 0.1$ and $0.5$.
(When the collisionality is much smaller than these values,
absolute linear instability is present \cite{usUniversalInstability}.)
It can be seen that increasing collisionality tends to limit amplification
at larger $k_y$ while having much less effect at smaller $k_y$.
This kinetic problem is similar to conventional hydrodynamics in that for both cases, non-modal amplification
decreases with increasing collisional dissipation (associated with viscosity in the Navier-Stokes equation
for the hydrodynamic case.)
However, the kinetic problem here is rather different
from the hydrodynamic case in the sensitivity of the transient amplification
to the geometric parameter $L_s/L_n$,
which is not associated with entropy-producing collisions.

@@

A variety of strategies can be used to initiate the nonlinear instability in numerical simulations.
Large-amplitude noise can be used, but this method is numerically inconvenient (at least in gs2) due to the rapid
drop in amplitude of strongly damped modes over the first few time steps,
which causes significant changes in the CFL-stable time step associated
with the perpendicular $\vect{E}\times\vect{B}$ speed,
thereby requiring recomputation of the implicit time-advance operator.
Another method for initiating the nonlinear instability is to begin with a sufficient
temperature gradient to bring about linear instability, and then to reduce the temperature gradient
once the turbulence reaches a sufficient amplitude \cite{Drake}.
A third effective method, the one used in figure \ref{fig:amplitudeDependence},
is to initialize with one of the least-damped linear eigenmodes scaled up to appropriate amplitude, with noise in the other modes.  We find this method to be particularly convenient since variation in the stable time step
is minimized.

>>>>

\newcommand{\red}[1]{\textcolor{red}{#1}}
\newcommand{\vect}[1]{\mbox{\boldmath $#1$}}
\newcommand{\erf}{\mathrm{erf}}
\newcommand{\PS}{Pfirsch-Schl\"{u}ter~}
\newcommand{\ExB}{\vect{E}\times\vect{B}}
\newcommand{\Cnl}{C_{\mathrm{nl}a}}
\newcommand{\vma}{\vect{v}_{\mathrm{m}a}}
\newcommand{\vm}{\vect{v}_{\mathrm{m}}}
\newcommand{\vda}{\vect{v}_{\mathrm{d}a}}
\newcommand{\vE}{\vect{v}_{E}}
\newcommand{\vdao}{\vect{v}_{\mathrm{d}a0}}
\newcommand{\vdo}{\vect{v}_{\mathrm{d}0}}
\newcommand{\vEo}{\vect{v}_{E0}}
\newcommand{\vpar}{v_{||}}
\newcommand{\vperp}{v_{\bot}}
\newcommand{\rhop}{\rho_{\theta}}
\newcommand{\rT}{r_{T}}
\newcommand{\Ti}{T_{\mathrm{i}}}
\newcommand{\rTi}{r_{T\mathrm{i}}}
\newcommand{\rTe}{r_{T\mathrm{e}}}
\newcommand{\reta}{r_{\eta}}
\newcommand{\rn}{r_{n}}
\newcommand{\rperp}{r_\bot}
\newcommand{\fM}{f_{\mathrm{M}}}
\newcommand{\fMa}{f_{\mathrm{M}a}}
\newcommand{\fMi}{f_{\mathrm{M}i}}
\newcommand{\fMe}{f_{\mathrm{Me}}}
\newcommand{\fe}{f_{\mathrm{e}}}
\newcommand{\Te}{T_{\mathrm{e}}}
\newcommand{\vth}{v_{\mathrm{th}}}
\newcommand{\nuii}{\nu_{ii}}
\newcommand{\nuPrime}{\hat{\nu}}
\newcommand{\gradpar}{\nabla_{||}}
\newcommand{\kpar}{k_{||}}
\newcommand{\psia}{\psi_0}
\newcommand{\psiaHat}{\hat{\psi}_0}
\newcommand{\omegase}{\omega_{*}}
\newcommand{\nuee}{\nu_{ee}}
\newcommand{\energy}{\mathcal{E}}
\newcommand{\sgn}{\mathrm{sgn}}
%\newcommand{\changed}[1]{\textcolor{red}{#1}}
\newcommand{\changed}[1]{#1}
\newcommand{\transpose}{\mathrm{T}}

@@

We consider slab geometry with magnetic shear, so the magnetic field in Cartesian
coordinates $(x',y',z')$ is $\vect{B} = [\vect{e}_{z'} + (x' / L_s) \vect{e}_{y'} ]B$
for some constant $B$ and shear length $L_s$.  A field-aligned coordinate system
is introduced: $x = x', \;y = y' - x' z' / L_s, \; z = z'$,
so $\vect{B}\cdot\nabla x = \vect{B}\cdot\nabla y = 0$ and $\vect{B}\cdot\nabla z = B$.
In this geometry, we consider the nonlinear electrostatic
gyrokinetic equation \cite{FriemanChen},
dropping toroidal effects:
\begin{eqnarray}
\label{eq:gke}
\frac{\partial h_s}{\partial t}
+ v_{||} \frac{\partial h_s}{\partial z}
+ \frac{1}{B} \left[
\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial X_s} \frac{\partial h_s}{\partial Y_s}
-\frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial Y_s} \frac{\partial h_s}{\partial X_s}
\right]
-\left< C_s\left\{h_s\right\} \right>_{\vect{R}_s} \\
=
\frac{q_s}{T_s} f_{Ms} \frac{\partial \left<\Phi\right>_{\vect{R}_s}}{\partial t}
-\frac{f_{Ms}}{B L_n} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right],
\nonumber
\end{eqnarray}
together with the quasineutrality condition
\begin{equation}
-\frac{n e \Phi}{T_i} + \int d^3v \left<h_i\right>_{\vect{r}}
=
\frac{n e \Phi}{T_e} + \int d^3v \left<h_e\right>_{\vect{r}}.
\label{eq:qn}
\end{equation}
Here $s \in \left\{i,e\right\}$ subscripts denote species,
$\Phi(t,\vect{r})$ is the electrostatic potential, $h_s(t,\vect{R}_s,\energy_s,\mu)$ is the nonadiabatic distribution function,
$\energy_s = m_s v^2/(2 T_s)$, $\mu = v_{\bot}^2/(2B)$,
and $f_{Ms}$ is the leading-order Maxwellian.
The scale length of the equilibrium density $n$ is $L_n = -n/(dn/dx)$,
$q_i = e = -q_e$ is the proton charge,
$T_s$ is the species temperature,
\changed{$\eta_s=(d \ln T_s/dx)/(d \ln n/dx)$,}
$C_s$ is the collision operator,
and
$\left<\right>_{\vect{r}}$
and
$\left<\right>_{\vect{R}_s}$
denote gyroaverages at fixed particle position $\vect{r}=(x,y,z)$ and guiding-center position $\vect{R}_s=(X_s,Y_s,Z_s)$.
The distinction between $z$ and $Z_s$ may be neglected.
\changed{
We consider a flux tube domain, in which the fluctuating quantities can be taken
to be periodic in $x$ and $y$.
}
Expanding the fluctuating quantities in spatial Fourier modes as $\Phi = \sum_{k_x,k_y} \Phi_{k_x,k_y}(t,z) \exp(i k_x x + i k_y y)$
and $h_s = \sum_{k_x,k_y} h_{s,k_x,k_y}(t,z,\energy_s,\mu) \exp(i k_x X_s + i k_y Y_s)$,
we may replace the gyro-averages with Bessel functions:
$\left<\Phi_{k_x,k_y} \exp(i k_x x + i k_y y)\right>_{\vect{R}_s}
=J_{0s} \Phi_{k_x,k_y} \exp(i k_x X_s + i k_y Y_s)$
and
$\left<h_{s,k_x,k_y}\exp(i k_x X_s + i k_y Y_s)\right>_{\vect{r}} = J_{0s} h_{s,k_x,k_y}\exp(i k_x x + i k_y y)$
with $J_{0s} = J_0(k_{\perp} v_{\perp} / \Omega_s)$
and $\Omega_s = e B/m_s$.
The perpendicular wavenumber is given by
\begin{equation}
k_{\perp} = \sqrt{k_y^2 + (k_x - k_y z/L_s)^2}.
\label{eq:kperp}
\end{equation}
The system is uniquely specified by the following dimensionless parameters:
$L_s / L_n$, $\eta_i$, $\eta_e$, $T_e / T_i$, and collisionality $\nuee L_n / v_i$,
where $\nuee = \sqrt{2}\pi n e^4 \ln\Lambda/[(4\pi\epsilon_0)^2 m_e^{1/2} T_e^{3/2}]$ is the electron-electron collision frequency
and $v_i = \sqrt{2 T_i / m_i}$.

@@

One important property of the system (\ref{eq:gke})-(\ref{eq:qn})
is the free energy conservation equation \cite{Howes, AbelReview}
obtained from the $\sum_s \int d^3R_s \int d^3v (T_s h_s/f_{Ms}) (\ldots)$ moment of (\ref{eq:gke}):
\begin{equation}
\frac{d W}{d t} = S(h_i,h_e),
\label{eq:dWdt}
\end{equation}
where
\begin{equation}
\label{eq:W}
W = \sum_s \int d^3r \int d^3v \frac{T_s \, \delta\! f_s^2}{2 f_{Ms}}
\end{equation}
is the free energy, $\delta\! f_s = h_s - q_s \Phi f_{Ms}/T_s$ is the total departure of the distribution
function from a Maxwellian,
and
\begin{eqnarray}
\label{eq:freeEnergySources}
S(h_i,h_e)
&=&
\sum_s \int d^3r \int d^3v \frac{T_s h_s}{f_{Ms}}  C_s\{ h_s \}  \\
&&+\sum_s
\int d^3R_s \int d^3v \frac{h_s}{B} \frac{\partial \left< \Phi\right>_{\vect{R}_s}}{\partial Y_s}
 \frac{T_s}{n_s}\frac{dn_s}{dx}
\left[ 1 + \left( \energy_s-\frac{3}{2}\right) \eta_s\right]
\nonumber
\end{eqnarray}
is the net energy source.
(To derive (\ref{eq:dWdt})-(\ref{eq:freeEnergySources}), the equalities
$\int d^3 R_s \int d^3v \left< \ldots\right>_{\vect{R}_s}
= \int d^3 R_s \int d^3v (\ldots)
\approx \int d^3 r \int d^3v (\ldots)
= \int d^3 r \int d^3v \left< \ldots\right>_{\vect{r}}
$ are used.
\changed{
The $v_{||} \partial h_s/\partial z$ term in (\ref{eq:gke}) has been integrated by parts;
as detailed on page 66 of \cite{BanonNavarroThesis}, the associated boundary term vanishes for a finite $z$ domain
using a twist-and-shift parallel boundary condition \cite{Beer1995},
taking the fluctuation amplitude to vanish at the ends of each set of linked domains.)}
The first term on the right-hand side of (\ref{eq:freeEnergySources})
is negative-definite (by the $H$-theorem).
The second term has the form of the particle and heat fluxes multiplied by
the density and temperature gradients, and this term increases $W$ if the fluxes have the appropriate sign
to relax the gradients.
Thus, in a turbulent steady state where $W$ is statistically constant,
equations (\ref{eq:dWdt})-(\ref{eq:freeEnergySources}) indicate
that energy is injected by the equilibrium gradient(s)
and dissipated by collisions.
Note that the operation $\int d^3R_s$ used in the derivation of
(\ref{eq:dWdt}) annihilates the nonlinear term
in (\ref{eq:gke}), and so (\ref{eq:dWdt})-(\ref{eq:freeEnergySources})
apply to both the linear and nonlinear dynamics.

@@

For the calculation in figure \ref{fig:transientAmplification}, the
initial condition used is the standard ``noise'' initial condition of
gs2: $g_s = h_s - \left< \Phi\right>_{\vect{R}_s} q_s f_{Ms} / T_s$ is
set to a Maxwellian velocity distribution times a random complex
number at each grid point in $z$.  (The real and imaginary parts of
these complex numbers are each taken from a uniform distribution on
$[-1,1]$.)  While techniques exist to calculate the optimal
perturbation that gives maximum amplification \cite{FarrellIoannou,
  SquirePRL, SquireApJ}, such techniques are numerically challenging
for this kinetic model due to the high dimensionality of the problem.
The noise initial condition is a convenient shortcut, ensuring energy
is likely to be given to a growing perturbation if one exists.  It has
been shown in \cite{Camargo, SquireApJ} that for studies of transient
linear amplification, random initial conditions are a reasonable proxy
for a directly calculated optimal initial condition.
\changed{Figure \ref{fig:transientAmplification}.b shows that 
when the calculation is repeated many times, the dynamics of the energy in figure \ref{fig:transientAmplification}.a are typical.}

Results for the three norms shown in figure
\ref{fig:transientAmplification}.a are quite similar, though the curve
for the $W$ norm appears slightly higher than the other curves due to
less decay in the initial period $t v_i / L_n <20$.  Indeed, we find
all results described or plotted for the remainder of this paper are
quite similar for all of these norms, 
\changed{using the locally-Maxwellian
initial perturbations described above.  However, there are certain
non-Maxwellian initial perturbations that can cause arbitrarily fast
growth in the electric field energy norms but not in the $W$ norm \cite{GregIan}.
Furthermore, it is $W$ that is conserved by the nonlinearities (recall
the conservation equation (\ref{eq:dWdt})) and thus the $W$ norm that
plays a role in determining bounds on the possibility of
nonlinearly-sustained turbulence, as we will show in section
\ref{sec:proof}.  For these reasons, we henceforth restrict our attention
to the $W$ norm.}

\begin{figure}[h!]
% Figure generated with m20141119_01_plotGS2TransientAmplificationForPaper_figureA.m
\includegraphics[width=3in]{m20141119_01_plotGS2TransientAmplificationForPaper_figureA_a.pdf}
\includegraphics[width=3in]{m20150305_02_plotManyIndntclGS2TransntAmplifctnRuns.pdf}
% Figure generated with m20141121_01_plotGS2TransientAmplificationForPaper.m
\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_c.pdf}
\includegraphics[width=3in]{m20141121_01_plotGS2TransientAmplificationForPaper_figuresBC_d.pdf}
\caption{(Color online) (a) Linear evolution for $L_s/L_n=300$, $\eta_i = \eta_e = 0$,
$\nuee = 0.05 v_i/L_n$, and $k_y \rho_i = 1$, illustrating transient amplification in three norms.
The initial conditions are detailed in the main text.
\changed{(b) Repeating the calculation 50 times, the results are always qualitatively similar.}
(c) The amount of transient amplification may be defined relative to $t=0$ or relative
to the minimum that occurs shortly thereafter.
(d) The potential structure along a field line for this example.
\label{fig:transientAmplification}}
\end{figure}

Figure \ref{fig:transientAmplification}.c shows the same computation, zooming in to the initial transient period.
Before the amplification begins around $t v_i/L_n=20$, there is an initial decay period due to strongly
damped modes. An amplification factor may be defined relative to either to this minimum amplitude,
or relative to the amplitude at $t=0$.
In both cases, we define the amplification factor to be 1
if there is no such transient amplification.

@@

\changed{
To demonstrate this principle across a range of parameters,
figure  \ref{fig:comparingTransientTo0Shear} demonstrates that the
correspondence between the finite-shear transients and zero-shear eigenmodes
is maintained throughout the evolution of the transient, for a range of
values of magetic shear.}
%This comparison is shown in figure \ref{fig:comparingTransientTo0Shear}.
In this figure, we display 25 independent simulations of linear transients,
varying $L_s/L_n$ between 20 and 300, and plotting the evolution of $\gamma = (2W(t))^{-1} dW/dt$ from the time
of peak growth rate to the time at which the energy begins to decay.
For all simulations shown, $k_y \rho_i = 2$ and $\nuee = 0.05 v_i/L_n$.
\changed{
Both $\left<|k_{||}|\right>$ and $\left< k_\perp \right>$ increase with time during the transient
growth, corresponding to movement upward and rightward in figure \ref{fig:localDispersionRelation}.a.
As a result, the effective zero-shear growth rate decreases over time.
}
For all transients we examined, there is very close agreement between the instantaneous growth rate of the transient
and the `predicted' growth rate from the zero-shear dispersion relation.
Thus, we can conclude that the transiently growing structures are closely related to the zero-shear
eigenmodes.  
%Magnetic shear causes $k_{\perp}$ and $k_{||}$ to vary with time as the structure
%evolves (unlike the zero-shear case in which $k_{\perp}$ and $k_{||}$ are constants.)

\begin{figure}[h!]
% Data for finite-shear transients were generated on Edison in
% /global/scratch2/sd/landrema/20141121-01-coderunner-edison-transientAmplificationScans/aky2/
%
% Data for zero-shear dispersion relation were generated on Edison in 
% /global/scratch2/sd/landrema/20150109-01-coderunner-edison-0shearUniversalInstability/vnewk_e_0p05
%
% Figure generated with m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.m
\includegraphics[width=4in]{m20150109_04_cmprngTransientGrowthRateTo0ShearForPaper.pdf}
\caption{(Color online)
Although a change from zero to small finite magnetic shear causes a fundamental change in the eigenvalues,
(which change from unstable to decaying),
the \emph{transient} finite-shear structures retain a similarity to the zero-shear modes.  For as shown here,
the instantaneous growth rate of the transient finite-shear structure is nearly identical
to the growth rate of a zero-shear eigenmode at the dominant $k_{\perp}$ and $k_{||}$ values
(\ref{eq:expectedKperp})-(\ref{eq:expectedKpar}).  Each trajectory shown is an independent
linear initial-value calculation, with $L_s/L_n$ varying from 20-300 between simulations.  The growth rate is plotted from
the time of maximum growth rate (at the $+$ symbols) until decay begins.
\label{fig:comparingTransientTo0Shear}}
\end{figure}

@@

\changed{ In contrast to the agreement shown in figure
  \ref{fig:comparingTransientTo0Shear} for the amplification phase
  following the maximum growth rate, we do not find such close
  agreement between the zero-shear dispersion relation and the
  finite-shear dynamics for very early times, before the maximum
  growth rate occurs.  In this initial period, a single dominant wave
  packet has not yet overtaken other fluctuations present in the
  initial random noise.  We also do not see such close agreement
  between the zero-shear and finite-shear behavior during the decay
  phase of the latter, i.e. at large times.  The finite-shear decay
  rate as $t \to \infty$ corresponds to an eigenvalue, and it was
  established in previous work such as \cite{Ross,Tsang,Antonsen} that
  the finite-shear eigenvalues are not continuously related to the
  zero-shear eigenvalues.  In an eigenmode, points at all $z$ have a
  coherent time-dependence.  Thus, this damping rate is determined by
  a coherent interaction between points of different
  $k_\perp(z)=k_y\sqrt{1+(z/L_s)^2}$, and between all $k_{||}$ Fourier
  components.  This damping rate is thus not well approximated by the
  zero-shear dispersion relation which assumes a single $k_\perp$ and
  single $k_{||}$. }

@@

Figure \ref{fig:transientAmplificationVsKy} shows the dependence
of the amplification factor on $k_y$ at fixed $L_s/L_n = 200$.
Significant amplification can be observed for a range of $k_y$ surrounding
$1/\rho_i$.
Results are shown for two values of collisionality,
$\nuee L_n / v_i = 0.1$ and $0.5$.
(When the collisionality is much smaller than these values,
absolute linear instability is present \cite{usUniversalInstability}.)
It can be seen that increasing collisionality tends to limit amplification
at larger $k_y$ while having much less effect at smaller $k_y$.
This kinetic problem is similar to conventional hydrodynamics in that for both cases, non-modal amplification
decreases with increasing collisional dissipation (associated with viscosity in the Navier-Stokes equation
for the hydrodynamic case.)

@@

A variety of strategies can be used to initiate the nonlinear instability in numerical simulations.
Large-amplitude noise can be used, but this method is numerically inconvenient (at least in gs2) due to the rapid
drop in amplitude of strongly damped modes over the first few time steps,
which causes significant changes in the CFL-stable time step associated
with the perpendicular $\vect{E}\times\vect{B}$ speed,
thereby requiring recomputation of the implicit time-advance operator.
Another method for initiating the nonlinear instability is to begin with a sufficient
temperature gradient to bring about linear instability, and then to reduce the temperature gradient
once the turbulence reaches a sufficient amplitude \cite{Drake}.
A third effective method, the one used in figure \ref{fig:amplitudeDependence},
is to initialize with one of the least-damped linear eigenmodes scaled up to appropriate amplitude, with noise in the other modes.  
We find this method to be particularly convenient since variation in the stable time step
is minimized.
\changed{
(In both simulations shown in figure  \ref{fig:amplitudeDependence}, the $k_y \rho_i=0$ Fourier component is initialized to 0,
but this component (and the $k_y \rho_i=\pm 0.2$ components) grow up quickly for the following reason. Since observable quantities are real, there is a large $k_y \rho_i=-0.1$
Fourier component present in the initial condition which is the complex conjugate of the large $k_y \rho_i=+0.1$ component shown,
and these two components interact nonlinearly to drive components at the sum and difference wave numbers.)
}
